---
title: "Functional programming in Data Science"
author: "Edwin Thoen"
date: "9/21/2017"
output: ioslides_presentation
---
---
title: "Functional Programming"
author: "Edwin"
date: "9/20/2017"
output: ioslides_presentation
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
set.seed(12345)
x <- data.frame(x1 = c(round(rnorm(3, 10, 2), 1), "NULL"),
                x2 = c(NA, round(runif(3, -10, 10), 1)),
                x3 = c(round(runif(3, -10, 10), 1), "NULL"),
                x4 = c(13.1, "null", 14.1, 8.3),
                stringsAsFactors = FALSE)
x_original <- x
```

## Intro

We have a dataset with three columns

```{r}
x
```

Columns `x1` and `x3` have a *NULL* or *null*. We need to fix this.

## Intro

Someone who just uses R or python would fix this maybe like

```{r}
x$x1[4] <- NA 
x$x3[4] <- NA 
x$x4[2] <- NA
```

## Intro

Someone who just uses R or python would fix this maybe like

```{r}
x$x1[4] <- NA 
x$x3[4] <- NA 
x$x4[2] <- NA
```

Then you realise that the three columns are strings because of the non-standard values.

```{r}
x$x1 <- as.numeric(x$x1)
x$x3 <- as.numeric(x$x3)
x$x4 <- as.numeric(x$x4)
```
There, fixed it!

## Intro

Downsides of fixing in place?

## Intro

Downsides of fixing in place?

> - **Not Scalable**

    Needs a new line of code for each new occurence. Infeasible for many occurences.
    
> - **Error Prone**

    Depends on the position of the character. Copy-paste errors.

```{r}
x$x3 <- as.numeric(x4)
```

> - **Code Overfit**

    Only works for this particular data set. Often we have renewal of data.
    
> - **


